---
title: "Peer Assessment1"
output: html_document
---

### Loading and preprocessing the data
```{r readdata,echo=TRUE}
activity <- read.csv('activity.csv', header=T, sep=",")
str(activity)
summary(activity)
```

### Bring some packages necessary for analysis
```{r librarypackages,echo=TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

### What is mean total number of steps taken per day?
#### 1. Calculate the total number of steps taken per day
```{r computenumberofsteps}
activity %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps)) %>%
    print(n=61)
```

#### 2. Make a histogram of the total number of steps taken each day.
```{r createhistogram}
data1 <- activity %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps)) 

hist(data1$tot_steps, main="Total Number of Steps per Day", 
     col="blue", xlab="Steps", ylab="Count (1,000)")
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day.
```{r calculatemean&median}
activity %>%
    group_by(date) %>%
    summarise(mean_steps = mean(steps, na.rm=TRUE),
              median_steps = median(steps, na.rm=TRUE)) %>%
    print(n=61)
```

### What is the average daily activity pattern?
#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval(x-axis) and the average
####    number of steps taken, averaged across all days (y-axis)
```{r timeseriesplot}
data2 <- activity %>%
    group_by(interval) %>%
    summarise(mean_steps = mean(steps, na.rm=TRUE))
data2 <- tbl_df(data2)
str(data2)
plot(data2$interval, data2$mean_steps, type="l", main="Time-series Plot",
     xlab="Interval", ylab="Average Steps")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum 
#### number of steps?
```{r maximumnumbersteps}
activity %>%
    group_by(interval) %>%
    summarise(max_step = max(mean(steps, na.rm = TRUE))) %>%
    summarise(himax_step = max(max_step, na.rm = TRUE)) 
```

### Imputing missing values
#### 1. Calculate and report the total number of missing values in the dataset 
#### (i.e. the total number of rows with NAs)
```{r calculatemissing}
colSums(is.na(activity))
```

#### 2. Devise a strategy for filling in all of the missing values in the dataset.
* Planning to use the mean substitution for missing values

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r meanofsteps}
activity2 <- transform(activity, steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))
activity2 <- tbl_df(activity2)
colSums(is.na(activity2))
summary(activity2)
```

#### 4.a. Make a histogram of the total number of steps taken each day.  
```{r histogram}
data_imputed <- activity2 %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps)) 

hist(data_imputed$tot_steps, main="Total Number of Steps per Day", 
     col="red", xlab="Steps", ylab="Count(1,000)")
```

#### 4.b. Calculate and report the mean and median total number of steps taken per day. 
```{r calculatemean} 
before_imp <- activity %>%
    group_by(date) %>%
    summarise(mean_steps = mean(steps, na.rm=TRUE),
              median_steps = median(steps, na.rm=TRUE)) 
    
after_imp <- activity2 %>%
    group_by(date) %>%
    summarise(mean_steps = mean(steps, na.rm=TRUE),
              median_steps = median(steps, na.rm=TRUE)) 
```

#### 4.c. Do these values differ from the estimates from the first part of the assignment? 
#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r summarycomparison}
summary(before_imp); summary(after_imp)

par(mfrow=c(1,2))
hist(data1$tot_steps, main="Total Number of Steps per Day", col="blue", 
     xlab="Steps", ylab="Count (1,000)")
hist(data_imputed$tot_steps, main="Total Number of Steps per Day", 
     col="red", xlab="Steps", ylab="Count(1,000)")
```

* There are more frequencies in the means for the data set with imputation than those for the orginal 
data set. Less variability is shown in the imputed data set.


### Are there differences in activity patterns between weekdays and weekends?
#### 1. Use the dataset with the filled-in missing values for this part.
#### Create a new factor variable in the dataset with two levels - 
#### "weekday" and "weekend" indicating whether a given date is a weekday or weekend day. 
```{r createfactorvar}
activity2$day <- as.Date(activity2$date)

weekdays_1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activity2$day_cat <- c('weekend', 'weekday')[(weekdays(activity2$day) %in% weekdays_1)+1L]
activity2$day_cat <- as.factor(activity2$day_cat)
summary(activity2$day_cat)
```


#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of 
#### the 5-minute interval (x-axis) and the average number of steps taken, 
#### averaged across all weekday days or weekend days (y-axis). 
```{r createpanelplot}
mean_steps <- mean(activity2$steps, na.rm=TRUE)

library(lattice)
xyplot(steps~interval | factor(day_cat), data=activity2, pch=19,
       main="Average Steps Taken for Interval", xlab="Interval", ylab="Avg Number of Steps",
       layout=c(1,2), type="l")
```


